FAIR fission track analysis with geochron@home

### 2025-09-28 geochron@home/manuscript/paper.tex
Updated Copernicus style sheets.

We here present the results of a type of `crowd-sourcing' experiment
that was carried out at UCL as part of an undergraduate course in
isotope geology. 70 students were asked to create an account on the
geochron@home server by selecting a unique username and
password. After spending a few minutes to complete the compulsory
tutorial (Section~\ref{sec:tutorial}), they were asked to count
fission tracks in randomly assigned image stacks of Mount Dromedary
apatite. Each student was required to analyse at least 15 grains,
using pre-defined ROIs. In a matter of hours, the students amassed a
dataset of 37,331 fission track counts in 34 grains.  This large
dataset was parsed into separate data files --- one for each student
--- which they processed during an assessed programming
exercise.\medskip

@@

The raw counts for 68 students (two of the students were unable to
complete the exercise) are given in Table~\ref{tab:crowdtable} and
Figure~\ref{fig:stripchart}, for the same selection of 25 grains that
were analysed by AC and PV in Section~\ref{sec:private}, along with
corresponding counts made by PV. Everyone counted tracks in the same
ROIs. Variation between the mean values is to be expected because of
the differing numbers of tracks due to varying areas and U contents of
the grains. But the variation between counts within grains (shown by
the standard deviations in Figure~\ref{fig:stripchart}) is due
entirely to differences between the students' recognising and counting
exactly the same tracks. These standard deviations increase with the
mean number of tracks and are considerable in size, being on average
about 40\% of the mean. With expert trained counters one would expect
much smaller differences between counts.  Furthermore, the vast
majority of students counted fewer tracks than PV did, often many
fewer, and on two occasions someone counted no tracks at all. PV's
count is always above the students' median and nearly always above the
upper quartile (Figure~\ref{fig:stripchart}).
Table~\ref{tab:crowdtable} (looking along rows) shows just 7 students
with average counts within 10\% of PV's, and 18 with average counts
less than half of his.\medskip

The rightmost three columns of Table~\ref{tab:crowdtable} contain
summary statistics ($n_g$ , $r_p$ and $s_p$) giving, for each student,
the number of grains counted and the weighted mean and standard
deviation of the ratios of the student's count to PV's count for each
grain (definitions provided in the caption). The true number of tracks
in each ROI is unknown, so we are comparing the students with PV
(remembering that PV's count is also subject to error, which is the
same for each student). $r_p$ measures agreement with PV on average
and $s_p$ measures consistency over grains. For good agreement $r_p$
should be close to 1 and $s_p$ should be small.  Perfect agreement is
$r_p = 1$ and $s_p = 0$. For comparison, an $r_p = 1.04$ and $s_p =
0.69$ for PV and AC using just the counts in the common areas $A_0$
from Section~\ref{sec:private}. With the caveat that the ROIs are
different and PV's counts are different, this calculation shows that
there is a group of 6 (possibly 9) students whose agreement with PV is
comparable with AC's. This group does not include student 229 whose
$r_p = 0.95$ but with a large $s_p = 1.62$.\medskip

However, in spite of the under-counting by the students, the ratio of
their pooled track counts ($r_p$) for two different grains is often
close to PV's ratio, especially when comparing grains with similar
track densities. This is illustrated in Figure~\ref{fig:23vs25}a for
grains~23 and 25.  For any point, the ratio of the two counts is given
by the slope of a line from the origin to that point. Although the
individual slopes vary considerably, the line with slope equal to the
ratio of the total counts for all students (which is 0.96) passes
close to PV's pair of counts (52,50) shown by the blue dot (with ratio
1.03). It is interesting that the students with lower counts are
mostly above this line and those with higher counts are all below it.
There is a positive correlation between the pairs of counts,
consistent with systematic observer effects (i.e., people who count a
low value in one grain tend to count similarly low in the other, and
vice versa). Comparing other pairs of grains shows similar results,
with correlations varying between 0.4 and 0.9.\medskip

>>>>

We here present the results of a type of `crowd-sourcing' experiment
that was carried out at UCL as part of an undergraduate course in
isotope geology. 68 students were asked to create an account on the
geochron@home server by selecting a unique username and
password. After spending a few minutes to complete the compulsory
tutorial (Section~\ref{sec:tutorial}), they were asked to count
fission tracks in randomly assigned image stacks of Mount Dromedary
apatite. Each student was required to analyse at least 15 grains,
using pre-defined ROIs. In a matter of hours, the students amassed a
dataset of 37,331 fission track counts in 34 grains.  This large
dataset was parsed into separate data files --- one for each student
--- which they processed during an assessed programming
exercise.\medskip

@@

The raw counts for the 68 students are given in
Table~\ref{tab:crowdtable} and Figure~\ref{fig:stripchart}, for the
same selection of 25 grains that were analysed by AC and PV in
Section~\ref{sec:private}, along with corresponding counts made by
PV. Everyone counted tracks in the same ROIs. Variation between the
mean values is to be expected because of the differing numbers of
tracks due to varying areas and U contents of the grains. But the
variation between counts within grains (shown by the standard
deviations in Figure~\ref{fig:stripchart}) is due entirely to
differences between the students' recognising and counting exactly the
same tracks. These standard deviations increase with the mean number
of tracks and are considerable in size, being on average about 40\% of
the mean. With expert trained counters one would expect much smaller
differences between counts.  Furthermore, the vast majority of
students counted fewer tracks than PV did, often many fewer, and on
two occasions someone counted no tracks at all. PV's count is always
above the students' median and nearly always above the upper quartile
(Figure~\ref{fig:stripchart}).\medskip

The rightmost three columns of Table~\ref{tab:crowdtable} contain
summary statistics ($n_g$, $r_p$ and $s_p$) giving, for each student,
the number of grains counted and the weighted mean and standard
deviation of the ratios of the student's count to PV's count for each
grain (definitions provided in the caption). The true number of tracks
in each ROI is unknown, so we are comparing the students with PV
(remembering that PV's count is also subject to error, which is the
same for each student). $r_p$ measures agreement with PV on average
and $s_p$ measures consistency over grains. For good agreement $r_p$
should be close to 1 and $s_p$ should be small.  Perfect agreement is
$r_p = 1$ and $s_p = 0$. For comparison, an $r_p = 1.04$ and $s_p =
0.69$ is obtained for PV and AC using just the counts in the common
areas $A_0$ from Section~\ref{sec:private}. With the caveat that the
ROIs are different and PV's counts are different, this calculation
shows that there is a group of 6 (possibly 9) students whose agreement
with PV is comparable with AC's. This group does not include student
229 whose $r_p = 0.95$ but with a large $s_p = 1.62$.\medskip

In spite of the under-counting by the students, the ratio of their
pooled track counts ($r_p$) for two different grains is often close to
PV's ratio, especially when comparing grains with similar track
densities. This is illustrated in Figure~\ref{fig:23vs25}a for
grains~23 and 25.  For any point, the ratio of the two counts is given
by the slope of a line from the origin to that point. Although the
individual slopes vary considerably, the line with slope equal to the
ratio of the total counts for all students (which is 0.96) passes
close to PV's pair of counts (52,50) shown by the blue dot (with ratio
1.03). It is interesting that the students with lower counts are
mostly above this line and those with higher counts are all below it.
There is a positive correlation between the pairs of counts,
consistent with systematic observer effects (i.e., people who count a
low value in one grain tend to count similarly low in the other, and
vice versa). Comparing other pairs of grains shows similar results,
with correlations varying between 0.4 and 0.9.\medskip
